\bf Model Visualization Techniques for a Social Network Model

### 2017-08-17 paper.Rnw
changes through sections 3.1.1

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. Yet another layout algorithm relies on the eigenstructure of the network's adjacency matrix to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes. 
\hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists} i\to j \\
                                   0 & otherwise
  \end{cases}
\]
In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.  
<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 4, arrowgap = .05, directed = T) +
  ThemeNet + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet + 
  theme(panel.grid = element_blank(), aspect.ratio = 1) +
  labs(title = "Adjacency Matrix Visualization") + 
  xlab("from (ego)") +
  ylab("to (alter)")
@
\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

Each type of visualization comes with its own advantages and disadvantages. Paths between nodes are easier to determine with node-link diagrams than with adjacency matrix visualizations. In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes. This is not necessarily possible in matrix representation, which is more focused on edges. Adjacency matrix visualization is particularly useful, however, when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $A$ twice: in $A_{ij}$ and $A_{ji}$.  \hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$. So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} \st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. Ultimately, t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper.

@@

A Stochastic Actor-Oriented Model (SAOM) is a model that is changing in time in order to accomodate for observations from the same network made at different points in time and that allows for changes in network structure due to actor-level covariates. This model was first introduced by Snijders in 1996 \citep{saompaper}. The two main properties of SAOMs, modelling the changes in time and using actor covariates, are crucial to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow, and most actors (or nodes) in social networks have inherent properties that could affect how they change their place within the network.

@@

In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process we usually condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), which we will refer to as $X(T)$. This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$. All other parts of the process are unseen. We model the process $X(T)$ as a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie. Once an actor is given the chance to change a tie, it tries to maximize a utility function based on the current and potential future states of the network.

>>>>

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet + 
  theme(panel.grid = element_blank(), aspect.ratio = 1) +
  labs(title = "Adjacency Matrix Visualization", x = "to (alter)", y = "from (ego)")
@
%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}. Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
%hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
Note here that our edge variables are binary: we only consider the presence or absence of an edge. If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.

Each type of visualization comes with its own advantages and disadvantages. For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges. Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.  \hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$. So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} \st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.

@@

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three components of dynamic networks: edge, node, and time information. It models the change of a network over time, allowing for changes in network structure due to actor-level covariates. This model was first introduced by Snijders in 1996 \citep{saompaper}. The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways, and most actors (or nodes) in social networks have inherent properties that could affect how they change their role within the network, and vice versa.

@@

In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process, we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC. The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize a utility function based on the current and potential future states of the network.
